Assessing primitives performance\\ on multi-stage execution

### 2018-05-13 paper.tex
oing over comparison

The Slang implementation is written inside the interpreter and has access to all interpreter APIs. Since the same primitive index is used for the primitive with 2 or 3 arguments, the primitive needs to deal with both cases in a way that the common case is efficient. The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write. This primitive follows the new specification, the full code is available in Appendix \ref{SlangAppendix}.

@@

In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end. The primitive is written in Cog's RTL, an abstract assembly which instructions compile almost one-to-one to native instructions. This primitive requires a more important amount of work to be written: the number of line of code is greater, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator. Since only performance critical implementations matter in this context, we've re-implemented only the case where there is 2 parameters. For the three parameter primitive, the VM falls-back to the Slang version. This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the readability of the approach, we use one of the most used metric, the number of lines of code.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails. This code is maintain by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} \cba{need to mention that this is for large strings}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead. \cba{InterpreterProxy not defined before}
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

>>>>

The Slang implementation is written inside the interpreter and has access to \sd{sd?full instead all} full interpreter APIs. Since the same primitive index \sd{no idea what is it} is used for the primitive with 2 or 3 arguments, the primitive needs to deal with both cases in a way that the common case is efficient. The primitive is now part of the core of the VM, it cannot be easily added or removed as plugin primitives, but it is simpler to read and write. This primitive follows the new specification, the full code is available in Appendix \ref{SlangAppendix}.

@@

In addition to the previous implementation, we've re-implemented here the primitive on top of Cog's JIT back-end. The primitive is written in Cog's RTL, an abstract assembly which \sd{whose?} instructions compile almost one-to-one to native instructions. This primitive requires a more important amount of work to be written: the number of line of code is larger, the implementor has to be careful to generate efficient machine code (there is no C compiler to optimize the code) and debugging such code requires the use of a processor simulator \sd{such as Bosch}. Since only performance critical implementations matter in this context, we've only re-implemented the case where there are two parameters. For the three parameter primitive, the VM falls back to the Slang version. This primitive follows the new specification, the full code is available in Appendix \ref{RTLAppendix}.

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). \sd{why not ST, Plugin, VM?}
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the approach readability, we use the number of lines of code even if there is no direct link between readability and mere lines of code. \sd{well this is kind of weak. but I do not know what do propose. Cyclomatic complexity counts the number of different exits. And it would be a nice complement}
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails. This code is maintain by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} \cba{need to mention that this is for large strings}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead. \cba{InterpreterProxy not defined before}
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}
